Static Analysis by Path Focusing

### 2011-05-03 report/report.tex
report

\part{Abstract Interpretation: state of the art}
\label{stateoftheartpart}
\section{Abstract Interpretation}

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. There have been different
work proposing a definition of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against the loss of precision induced by widening.

@@

\end{tikzpicture}
   \end{minipage}
   \caption{Example of program, and its associate control flow graph. This
   program comes from \cite{GopanR06}.}
\end{figure}
\FloatBarrier

@@

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to different reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, one compute a union of convex polyhedra instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

@@

@@

	Path focusing technique uses control flow graphs in static single assignment
	(SSA) form. The main concept behind the SSA form is that every variable is
	only being assigned once.

@@

	This is a reachability problem, that can be expressed as an SMT-formula,
	which is a logic formula with boolean variables, and elements of a certain
	theory $T$. Depending on the program, one can choose different theories:
	\begin{itemize}
	\item For programs with rational variables, whose operations (instructions,
	transition conditions\dots) are all linear arithmetic, $T$ can be the theory
	of linear real arithmetic (LRA).
	\item For programs with integer variables (with a numerical state space in
	$\Z^n$) $T$ can be the theory of linear integer arithmetic (LIA).
	\end{itemize}

	Despite deciding the satisfiability of such formula is NP-complete, there
	has been lots of research on decision procedures \cite{Kroening08} and there exist nowadays
	efficient programs, known as SMT-solvers, that can decide the satisfiability
	of LRA or LIA formulas. Well-known efficient SMT-solvers are
	Z3 \cite{MouraB08} and Yices \cite{DutertreM06}.

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the boolean variables such that the formula is true.
	Among these boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
\begin{algorithm}
\caption{Path Focusing}\label{pathfocusingalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\State $Y \gets \tau_e^\#(X_{p_i})$
		\If {$p_j \in P_W$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly perform some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	To Be Completed\dots

@@

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
compute this number of path at the beginning of the algorithm.

\part{Implementation}\label{implementationpart}

@@

The advantage of using such an infrastructure is that the analyzer can check
programs written in various languages: C, C++, \dots, without needing to write
a specific frontend for each of them. Here, the analyzer works directly on the
internal representation (IR), which is a graph representation of
the program. The control flow graph can easily be extracted from
this representation.

@@

During the implementation, we encountered problems due to the use of the LLVM
internal representation. Indeed, this format is not provided for static
analysis, but for code generation. Then, some informations we need to do precise
static analysis are lacking.

@@

@@

		The analyzer is implemented with the APRON library \cite{JM09}, which
		is a common interface to libraries implementing all the features
		for several abstract domains, such as convex polyhedra (NewPolka), 
		octagons (OCT), intervals (BOX).
		In our case, we used the convex polyhedra library, but we could switch
		to octagons or intervals easily.

		\subsection{Dimensions of the Abstract values}
		To compute the set of live variables in SSA: \cite{Boi10}

>>>>

\part{Abstract Interpretation: state of the art}
\label{stateoftheartpart}

\section{Introductive example}

We consider the following program:

\begin{figure}[!h]
   \begin{minipage}[c]{.46\linewidth}
\begin{C}
x = 0;
while (x < 100) {
	x++;
}
\end{C}
   \end{minipage} \hfill
   \begin{minipage}[c]{.46\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.7cm,
                    semithick,font=\footnotesize]

	\node[state] (n0) {$p_0$};
	\node[state] (n1) [below of=n0] {$p_1$};
	\node[state] (n2) [below of=n1] {$p_2$};
	\node[state] (n3) [left of=n1] {$p_3$};

  \path [transition] 
		(n0) edge              node {$x = 0$} (n1);
  \path [transition] 
        (n1) edge			   node [above] {$x \geq 100?$} (n3);
  \path [transition] 
        (n1) edge			   node [left] {$x < 100?$} (n2);
  \path [transition] 
        (n2) edge [out=0, in=0, distance=2cm] node [right] {$x = x+1$} (n1);

\end{tikzpicture}
   \end{minipage}
\end{figure}
\FloatBarrier

Static analysis aims to discover some properties about programs. In this
example, one would to compute the set of possible values for the variable $x$
during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}

We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more elements in the sets\dots
\end{itemize}

This computation is a fixpoint computation: one updates the different sets until
we find no more elements. These sets are strictly increasing (we always add new
elements in the sets), and if we find no new elements to add, that means we have
reached a limit.

\section{Abstract Interpretation}

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. There have been various
work proposing a definition of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against the loss of precision induced by widening.

@@

\end{tikzpicture}
   \end{minipage}
   \caption{Example of program, and its associate control flow graph. This
   program comes from \cite{GopanR06}.}
\label{runningexample}
\end{figure}
\FloatBarrier

@@

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to various reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, one compute a union of convex polyhedra instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

@@

In some cases, a loop can have a non-regular behaviour, especially when it has
several paths. Some of these paths can stay impossible for a while, and become
possible after a certain number of iterations. However, the widening operator
extrapolates without taking care of these new possible paths. This can induce a
loss of precision, that Lookahead widening aims to limit.

Lookahead widening \cite{GopanR06} is a technique that aims to isolate the
different loop phases during the analysis, in order to obtain more precise
results.

In the program already depicted in figure \ref{runningexample}, there is a loop
with two perfectly distinct phases:
\begin{itemize}
\item During the 51 first iterations, both $x$ and $y$ are incremented.
\item During the 51 last iterations, $x$ is incremented and $y$ is decremented.
\end{itemize}

When computing the fixpoint iterations, we only consider a subset of the graph
for a while, until the convergence. In our example:
\begin{itemize}
\item Step 1:
	we only consider the path of
	the loop that is possible at the first iteration. The second path is
	ignored, and will be treated later. we compute the iterations, and we obtain
	an first invariant for this part of the graph. Then, we can do some
	narrowing steps in order to recover some precision, before adding new paths
	in the graph.
\item Step 2:
	Now, we add the new paths into our graph if they have become possible. In
	our case, the second path of the loop is now possible, so we add the path in
	the graph and compute new iterations until convergence\dots
\item Step 3: finally, the path from $p_4$ to $p_7$, so we add it into the graph and
compute the invariant for $p_7$.
\end{itemize}

\begin{figure}[!h]
   \begin{minipage}[c]{.46\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
                    semithick,font=\footnotesize]

  \path [transition] 
		(n0) edge              node {$x,y \leftarrow 0$} (n1);
  \path [transition] 
        (n1) edge			   node [left] {$x \leq 50$} (n2);
  \path [transition2] 
        (n1)  edge              node [right] {$x \geq 51$} (n3);
  \path [transition] 
        (n2) edge              node [left] {$y \leftarrow y+1$} (n4);
  \path [transition2] 
        (n3) edge			   node [right] {$y \leftarrow y-1$} (n4);
  \path [transition] 
        (n4) edge			   node {$y \geq 0$} (n5);
  \path [transition2] 
		(n4) edge  [out = 180, in=90] node [left] {$y \leq -1$} (n7);
  \path [transition] 
        (n5) edge              node {$x \leftarrow x+1$} (n6);
  \path [transition] 
        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);

\end{tikzpicture}
\centering Step 1
   \end{minipage} \hfill
   \begin{minipage}[c]{.46\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
                    semithick,font=\footnotesize]

  \path [transition] 
		(n0) edge              node {$x,y \leftarrow 0$} (n1);
  \path [transition] 
        (n1) edge			   node [left] {$x \leq 50$} (n2);
  \path [transition] 
        (n1)  edge              node [right] {$x \geq 51$} (n3);
  \path [transition] 
        (n2) edge              node [left] {$y \leftarrow y+1$} (n4);
  \path [transition] 
        (n3) edge			   node [right] {$y \leftarrow y-1$} (n4);
  \path [transition] 
        (n4) edge			   node {$y \geq 0$} (n5);
  \path [transition2] 
		(n4) edge  [out = 180, in=90] node [left] {$y \leq -1$} (n7);
  \path [transition] 
        (n5) edge              node {$x \leftarrow x+1$} (n6);
  \path [transition] 
        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);

\end{tikzpicture}
\centering Step 2
   \end{minipage}
   \caption{Dotted arrows are ignored by the analysis.}
\label{gopanreps}
\end{figure}
\FloatBarrier

@@

	Path focusing technique uses control flow graphs in static single assignment
	(SSA) form. The main concept behind the SSA form is that, syntactically,
	every variable is	
	only being assigned once.

@@

	This is a reachability problem, that can be expressed as an SMT-formula,
	which is a logic formula with Boolean variables, and elements of a certain
	theory $T$. Depending on the program, one can choose different theories:
	\begin{itemize}
	\item For programs with rational variables, whose operations (instructions,
	transition conditions\dots) are all linear arithmetic, $T$ can be the theory
	of linear real arithmetic (LRA).
	\item For programs with integer variables (with a numerical state space in
	$\Z^n$) $T$ can be the theory of linear integer arithmetic (LIA).
	\end{itemize}

	Although deciding the satisfiability of such formula is NP-complete, there
	has been much research on decision procedures \cite{Kroening08} and there exist nowadays
	efficient programs, known as SMT-solvers, that can decide the satisfiability
	of LRA or LIA formulas. Well-known efficient SMT-solvers are
	Z3 \cite{MouraB08} and Yices \cite{DutertreM06}.

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the Boolean variables such that the formula is true.
	Among these Boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the Boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
	We maintain a set $A$ of point in $P_R$ that need to be treated.
	We select iteratively one of these control points, and search for new focus
	paths starting from this point, until there is no more path to be treated.
	At the end of the analysis, one could perform some narrowing iterations in
	order to recover some precision.
\begin{algorithm}
\caption{Path Focusing}\label{pathfocusingalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\State $Y \gets \tau_e^\#(X_{p_i})$
		\If {$p_j \in P_W$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly perform some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply apply the widening
	operator, one could compute some narrowing steps to regain precision before
	focusing on another path. This is the same principle as the one we find in
	\emph{Lookahead widening} \cite{GopanR06}.

	So, the algorithm has a special case if the path we focus on goes from $p_i$
	to itself: instead of computing $\tau_e^\#(X_{p_i})$, one compute
	$SelfLoop(\tau_e^\#,X_{p_i})$, defined in algorithm \ref{loopiteralgo}.

	The $SelfLoop$ procedure actually computes a widening operation, and
	computes a narrowing step in order to recover precision. In some cases,
	instead of performing widening/narrowing iterations, one could use
	acceleration techniques to find even more precise invariants \cite{GH06}.

	In order for this refinement to be precise, one needs to delay the widening
	operator when we consider a self-loop for the first time. Then, our
	algorithm uses a set $U$ of paths that have already been treated. If the
	self-loop we focus on is not in $U$, we apply a simple union instead of
	widening. Widening will be applied next time to guarantee the convergence.

\begin{algorithm}
\caption{SelfLoop}
\label{loopiteralgo}
\begin{algorithmic}[1] 
\Procedure{SelfLoop}{$\tau_e^\#$,$X_{p_i}$}
	\State $Y \gets \tau_e^\#(X_{p_i})$
	\State $X' \gets X_{p_i} \sqcup Y$
	\State $Y \gets \tau_e^\#(X')$
	\State \textbf{return} $Y$
\EndProcedure
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Path Focusing with special treatment for self loops}
\label{pathfocusingoptalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\State $U \gets \emptyset$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\If{$p_i = p_j$}
			\State $Y \gets SelfLoop(\tau_e^\#,X_{p_i})$
		\Else
			\State $Y \gets \tau_e^\#(X_{p_i})$
		\EndIf
		\If {$(p_j \in P_W) \wedge (p_i \neq p_j \vee e \in U)$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
			\State $U \gets U \cup \{e\}$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly perform some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
compute this number of paths at the beginning of the algorithm.

\FloatBarrier
\part{Implementation}\label{implementationpart}

@@

The advantage of using such an infrastructure is that the analyzer can check
programs written in various languages: C, C++, \dots, without needing to write
a specific frontend for each of them. Here, the analyzer directly works on the
internal representation (IR), which is a graph representation of
the program. The control flow graph can easily be extracted from
this representation.

Our analyzer takes as parameter the LLVM internal representation of a program,
and returns linear relations between the numerical variables at each control
point of this program. It first applies some optimisation passes over the IR,
and then computes the linear relations by abstract interpretation.

@@

During the implementation, we encountered problems due to the use of the LLVM
internal representation. Indeed, this format is not provided for static
analysis, but for code generation. Then, some informations we need for a precise
static analysis are lacking.

@@

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed after the register
allocation. Then, the data structure we work on doesn't have live variables
informations yet. To solve this problem, we coded our own pass, that aims to
compute these live variables after the transformation in SSA form.

@@

	The analyzer is implemented using the APRON library \cite{JM09}, which
	is a common interface to libraries implementing all the features
	for several abstract domains, such as convex polyhedra (NewPolka), 
	octagons (OCT), intervals (BOX).
	In our case, we used the convex polyhedra library, but we could switch
	to octagons or intervals easily.

	\subsection{Dimensions of the Abstract values}

	In our implementation, we tried to reduce as much as possible the dimensions
	of the abstract values at each program point. Indeed, there is no need to
	consider the whole set of the program numerical variables at each point $p$,
	since lots of these variables are not \emph{live} in $p$. 
	The definition of a \emph{live} variable is the following:

	\begin{definition}[Liveness]
	A variable is \emph{live} at the control point $p$ iff:
	\begin{itemize}
	\item its value is available at $p$, meaning that there is a path from the
	point defining the variable and $p$.
	\item its value might be used in the future, meaning that there exist a path
	starting in $p$, that arrives in a point using the variable.
	\end{itemize}
	\end{definition}

	LLVM internal representation allows to find the control point where
	a variable is defined and used, thanks to def-use chains: we can
	easily get the set of points where a variable is used, and we directly have
	a pointer to its definition.
	With all these informations, we implemented an LLVM pass that computes 
	the set of live variables in SSA, using the liveness check algorithm
	described in \cite[section 3.3]{Boi10}.

	At a program point $p$, the dimensions of the associate abstract value could
	only be the numerical variables that are live at $p$. This is
	not exactly the case in our implementation, because of another 
	optimisation: we only consider
	variables that are not a linear combination of other variables.

	For instance, assume that $x,y,z$ are numerical variables of a program,
	$x$ is defined as $x = y+z$, and $x,y,z$ are live at point $p$. Instead of having
	$x$ as a dimension for $X_p$, we only have $y$ and $z$. All the properties
	for $x$ can be directly extracted from $X_p$ and the information $x=y+z$.
	This is an optimisation in the sense that there is redundant information in
	the abstract value if both $x,y$ and $z$ are dimensions of $X_p$.

	Then, liveness definition can be adapted in our case:

	\begin{definition}[Liveness\dots]
	A variable $v$ is \emph{live\dots} at the control point $p$ iff:
	\begin{itemize}
	\item 
	its value is available at $p$, meaning that there is a path from the
	point defining the variable to $p$,
	\item one of these conditions holds:
		\begin{itemize}
		\item $v$ is live in $p$.
		\item There is a variable $v'$, defined as a linear combination of other
		variables $v_1, v_2, \dots, v_n$, such that $\exists i \in \{1,\dots,n\}, v = v_i$,
		and $v'$ is live\dots in $p$.
		\end{itemize}
	\end{itemize}
	\end{definition}

	Finally, a variable is a dimension of $X_p$ iff it is live\dots and it is
	not defined as a linear combination of program variables.

	\subsection{SMT-solving}

	\emph{Path focusing} technique requires to decide the satisfiability of some
	SMT formula, and to extract a model when the formula is satisfiable, i.e an
	assignement of the variables such that the formula is \emph{true}. For doing
	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
	and Yices \cite{DutertreM06}. One can easily switch from one to another with
	an argument we give as parameter to the analyzer. In this way, we can
	compare the effectiveness of both SMT-solvers.
